---
title: "UCSL R Core Challenge"
author: "Christopher Prince (cmp670@nyu.edu)"
date: "08/02/2015"
output: 
  pdf_document:
    number_sections: true
---

```{r}
fname <- '~/UCSL/R/ch2/2014-07 - Citi Bike trip data.csv'
#fname <- '~/UCSL/R/ch2/Citi Bike Clean Data.csv'   #smaller set for testing
cbd<-read.csv(fname,header=TRUE)
cbd<-data.frame(cbd, byear = as.numeric(as.character(cbd$birth.year))) #probably a better way to do this...
```

#Setup
In addition to calculating the z-score as in previous challenges, we also establish a naive distance metric (great circle distance between start and end, not distance travelled).
```{r include=FALSE}
#get tripduration parameters for z-score calcluation
tdur.mean <- mean(cbd$tripduration)
tdur.sd <- sd(cbd$tripduration)
#Create a z-score based on tripduration
z_sc <- (cbd$tripduration-tdur.mean)/tdur.sd
```

```{r}
distance <- function(slat, slon, elat, elon){
  return acos(sin(slat*pi/180)*sin(elat*pi/180)+cos(slat*pi/180)*cos(elat*pi/180)*cos((slon-elon)*pi/180))*637100
}
dist <- distance(cbd$start.station.latitude, cbd$start.station.longitude, cbd$end.station.latitude, cbd$end.station.longitude)
```
```{r include=FALSE}
#append these to our data frame
cbd <- cbind(cbd, dist, z_sc)
```

#Descriptive statistics
##Summaries
```{r}
##no outlier removal
#Data summaries
summary(cbd$tripduration)
summary(cbd$byear)
summary(cbd$dist)
```

##Histograms
```{r}
par(mfrow = c(3,1))
hist(cbd$tripduration)
hist(cbd$byear)
hist(cbd$dist)
```

##Boxplots
```{r}
par(mfrow = c(1,3))
boxplot(cbd$tripduration)
boxplot(cbd$byear)
boxplot(cbd$dist)
```

##Scatter plots
These take a long time to render, so we note in passing that a simple call to `plot` the variables results in a sparse graph that is difficult to read. We include here only a version with axes cutoff to a region of interest and using very small symbols.
```{r}
par(mfrow = c(1,1))
#plot(cbd$byear, cbd$tripduration)
symbols(cbd$byear, cbd$tripduration, circles=array(0,dim(cbd)[1]), xlim=c(1920,2000), ylim=c(60,2000))
```

##Intensity plots
Even after optimizing the scatter plot above, it is difficult to see how the density of these plots vary with nearly one million data points, so we construct intensity plots by creating histograms across `byear`.

```{r}
td1<-array(0,c(81,74))

count=0
for (i in c(1920:2000)){
  count<- count + 1
  s<-subset(cbd, byear==i)
  t<-hist(s$tripduration, breaks=60*c(1:60,15*c(5:12),60*c(4:10)), plot=FALSE)
  td1[count,]=t$counts
}

image(c(1920:2000), 60*c(1:60,15*c(5:12),60*c(4:10)), td1, xlim=c(1920,2000), ylim=c(60,2000), col=rainbow(30), xlab='Birth year', ylab='Trip duration')
title(main = 'Relative intensity of rides taken by trip duration and birth year')

td2<-array(0,c(81,40))

count=0
for (i in c(1920:2000)){
  count<- count + 1
  s<-subset(cbd, byear==i)
  t<-hist(s$dist, breaks=250*c(0:40), plot=FALSE)
  td2[count,]=t$counts
}

image(c(1920:2000), 250*c(1:40), td2, xlim=c(1920,2000), ylim=c(0,6000), col=rainbow(30), xlab= 'Birth year', ylab='Distance')
title(main = 'Relative intensity of rides taken by distance and birth year')
```

#Regression analysis
##Function setup
```{r}
#Create a regression function to reuse in this document
myregression <- function(x, y){
  x.mean <- mean(x, na.rm = TRUE)
  y.mean <- mean(y, na.rm = TRUE)
  
  #Data frame to calculate regression manually
  data <- data.frame(x, y, xdev=(x-x.mean), ydev=(y-y.mean), xdevydev=((x-x.mean)*(y-y.mean)), xdev2=(x-x.mean)^2,ydev2=(y-y.mean)^2)
  
  #OLS terms
  SP <- sum(data$xdevydev, na.rm=TRUE)
  SSx <- sum(data$xdev2, na.rm=TRUE)
  SSy <- sum(data$ydev2, na.rm=TRUE)
  SSxy <- sum(data$xdevydev, na.rm=TRUE)
  b1 <- SP / SSx
  b0 <- y.mean-b1*x.mean
  
  #Manual calculation of R-squared
  r2 <- SSxy^2/(SSx*SSy)
  
  #get the set names
  xname <- deparse(substitute(x))
  yname <- deparse(substitute(y))
  
  #return the results
  results <- list(b0=b0,b1=b1,r2=r2,xmean=x.mean,ymean=y.mean,xname=xname, yname=yname, x=x,y=y) #, model=model1)
  
  return(results)
}

#plot the regression on an xy plot
myRegPlots <- function(myregr, xlim=c(min(myregr$x, na.rm=TRUE),max(myregr$x, na.rm=TRUE)), ylim=c(min(myregr$y, na.rm=TRUE),max(myregr$y, na.rm=TRUE))){
  plot(myregr$x,myregr$y,col="red", lwd = 1, xlim=xlim, ylim=ylim, xlab=myregr$xname, ylab=myregr$yname, xaxs='i')
  points(myregr$xmean,myregr$ymean,col="green", lwd = 10)
  lines(xlim,myregr$b0+myregr$b1*xlim, lwd=4)  
}
```

##Models
###Tripduration as a function of birth year:
```{r}
modelBYxTDUR<-myregression(x = cbd$byear, y = cbd$tripduration)
myRegPlots(myregr = modelBYxTDUR,ylim = c(0,2000))

image(c(1920:2000), 60*c(1:60,15*c(5:12),60*c(4:10)), td1, xlim=c(1920,2000), ylim=c(60,2000), col=rainbow(30))
lines(c(1920:2000),modelBYxTDUR$b0+modelBYxTDUR$b1*c(1920:2000), lwd=3)  
points(modelBYxTDUR$xmean,modelBYxTDUR$ymean,col="black", lwd = 3, bg="white", pch=21)

#Verify using R's lm() method
summary(lm(cbd$tripduration~cbd$byear))
```

###Distance as a function of birth year:
```{r}
modelBYxDIST<-myregression(x = cbd$byear, y = cbd$dist)
myRegPlots(myregr= modelBYxDIST)
image(c(1920:2000), 250*c(1:40), td2, xlim=c(1920,2000), ylim=c(60,6000), col=rainbow(30))
lines(c(1920:2000),modelBYxDIST$b0+modelBYxDIST$b1*c(1920:2000), lwd=3)  
points(modelBYxDIST$xmean,modelBYxDIST$ymean,col="black", lwd = 3, bg="white", pch=21)
summary(lm(cbd$dist~cbd$byear))

#Subset data to remove outliers (z_sc<2)
cbd.z2<-subset(cbd, z_sc<2)

##Outlier removal
#Data summaries
summary(cbd.z2$tripduration)
summary(cbd.z2$byear)
summary(cbd.z2$dist)

#Histograms
par(mfrow = c(3,1))
hist(cbd.z2$tripduration)
hist(cbd.z2$byear)
hist(cbd.z2$dist)

#Boxplots
par(mfrow = c(1,3))
boxplot(cbd.z2$tripduration)
boxplot(cbd.z2$byear)
boxplot(cbd.z2$dist)

#xy-plots
par(mfrow = c(1,1))
plot(cbd.z2$byear, cbd.z2$tripduration)
symbols(cbd.z2$byear, cbd.z2$tripduration, circles=array(0,dim(cbd.z2)[1]), xlim=c(1920,2000))
symbols(cbd.z2$byear, cbd.z2$tripduration, circles=array(0,dim(cbd.z2)[1]), xlim=c(1920,2000), ylim=c(60,2000))

#intensity plots
td3<-array(0,c(81,74))

count=0
for (i in c(1920:2000)){
  count<- count + 1
  s<-subset(cbd.z2, byear==i)
  t<-hist(s$tripduration, breaks=60*c(1:60,15*c(5:12),60*c(4:10)), plot=FALSE)
  td3[count,]=t$counts
}

image(c(1920:2000), 60*c(1:60,15*c(5:12),60*c(4:10)), td3, xlim=c(1920,2000), ylim=c(60,2000), col=rainbow(30))
#image(c(1920:2000), 60*c(1:60,15*c(5:12),60*c(4:10)), td3, xlim=c(1920,2000), ylim=c(60,2000), col=c(gray(100:0/100)))

td4<-array(0,c(81,40))

count=0
for (i in c(1920:2000)){
  count<- count + 1
  s<-subset(cbd.z2, byear==i)
  t<-hist(s$dist, breaks=250*c(0:40), plot=FALSE)
  td4[count,]=t$counts
}

image(c(1920:2000), 250*c(1:40), td4, xlim=c(1920,2000), ylim=c(60,6000), col=rainbow(30))
#image(c(1920:2000), 250*c(1:40), td4, xlim=c(1920,2000), ylim=c(60,6000), col=c(gray(100:0/100)))

##Models
#Start with tripduration as a function of birth year:
modelBYxTDUR2<-myregression(cbd.z2$byear, cbd.z2$tripduration)
myRegPlots(modelBYxTDUR2)
image(c(1920:2000), 60*c(1:60,15*c(5:12),60*c(4:10)), td3, xlim=c(1920,2000), ylim=c(60,2000), col=rainbow(30))
lines(c(1920:2000),modelBYxTDUR2$b0+modelBYxTDUR2$b1*c(1920:2000), lwd=3)  
points(modelBYxTDUR2$xmean,modelBYxTDUR2$ymean,col="black", lwd = 3, bg="white", pch=21)
summary(lm(cbd.z2$tripduration~cbd.z2$byear))

#Repeat calculations with our distance metric as a function of birth year:
modelBYxDIST2<-myregression(cbd.z2$byear, cbd.z2$dist)
myRegPlots(modelBYxDIST2)
image(c(1920:2000), 250*c(1:40), td4, xlim=c(1920,2000), ylim=c(60,6000), col=rainbow(30))
lines(c(1920:2000),modelBYxDIST2$b0+modelBYxDIST2$b1*c(1920:2000), lwd=3)  
points(modelBYxDIST2$xmean,modelBYxDIST2$ymean,col="black", lwd = 3, bg="white", pch=21)
summary(lm(cbd.z2$dist~cbd.z2$byear))

#Repeat calculations with our distance metric as a function of birth year:
modelTDURxDIST2<-myregression(cbd.z2$tripduration, cbd.z2$dist)
myRegPlots(modelTDURxDIST2)

td5<-array(0,c(40,74))

count=0
for (i in 250*c(1:40)){
  count<- count + 1
  s<-subset(cbd.z2, abs(dist-i)<=250)
  t<-hist(s$tripduration, breaks=60*c(1:60,15*c(5:12),60*c(4:10)), plot=FALSE)
  td5[count,]=t$counts
}
image(60*c(1:60,15*c(5:12),60*c(4:10)), 250*c(1:40), t(td5), xlim=c(0,2500), ylim=c(0,10000), col=rainbow(30))
lines(50*c(0:50),modelTDURxDIST2$b0+modelTDURxDIST2$b1*50*c(0:50), lwd=3)  
points(modelTDURxDIST2$xmean,modelTDURxDIST2$ymean,col="black", lwd = 3, bg="white", pch=21)

summary(lm(cbd.z2$tripduration~cbd.z2$dist))
```
