---
title: "UCSL: R Challenge 2"
author: "Christopher Prince (cmp670@nyu.edu)"
date: "07/22/2015"
output: 
  pdf_document:
    number_sections: true
---

```{r setup, echo=FALSE}
#set display options for output values in document
options('scipen'=2, 'digits' = 2)
#library for prettyprinting quartile tables with kable()
library(knitr)
```

# *The majority of the trips are for short commutes lasting no more than 15min.*
## Test plan
We first state the null and alternative hypotheses. Denoting the population median with $\tilde{\mu}$,
$$H_0: \tilde{\mu} \leq 900$$
$$H_a: \tilde{\mu} > 900$$

Because the alternative hypothesis is a "greater than" condition, we need to use an upper-tail hypothesis test. For a t-test, this means that we will reject the null hypothesis if the statistic is greater than the critical value for the test at the desired confidence. We will assume $\alpha = 0.05$.

## Importing and summarizing the data
```{r}
#import Jan 2015 'clean' CitiBike data from csv file, include a header line
fname <- '~/UCSL/R/ch2//Citi Bike Clean Data.csv'
cbdata <- read.csv(fname, header = TRUE)
```

```{r}
#Print the summary data and calculate the sample parameters n, s, xbar and xmed:
summary(cbdata$tripduration)
n <- dim(cbdata)[1]
s <- sd(cbdata$tripduration)
xbar <- mean(cbdata$tripduration)
xmed <- median(cbdata$tripduration)
```

```{r}
#Fit a normal distribution to the data. Overlay a normal distribution curve onto the histogram of the data.
par(mfrow = c(1,2))
hist(cbdata$tripduration, breaks = c((1:60)*60,Inf), probability = TRUE, xlim=c(0,3600))
xfit <- seq(xbar-4*s,xbar+4*s)
yfit <- dnorm(xfit, mean = xbar, sd = s)
lines(xfit, yfit, lwd=3, col='orange')
boxplot(cbdata$tripduration, ylim=c(0,3600))
```


Before we even do any analysis, we note that the null hypothesis is heavily favored in this sampling, since 68.2% of values are less than 900. It would be remarkable for a sample of over 20,000 observations be so off-balance from a true median over 900. Nevertheless, we perform the calculations to support this.

Because $n=20853$, and the histogram is not wildly skewed (out-of-normal), CLT applies. But how do we account for the fact that the the t-test hypotheses are not on means? CLT tells us that means of a sufficiently large number of samples from a population of any distribution will themselves distribute normally. Does this apply to medians as well? Intuitively, subsampled medians seem like they should distrubute normally in the same manner that the mean does. Let's run an experiment. First, the histogram suggests that this may follow something like a log-normal distribution. The histogram of the log of the `tripduration`s looks like so:

```{r}
xbarlog<-mean(log(cbdata$tripduration))
slog<- sd(log(cbdata$tripduration))
xfitlog<-seq(xbarlog-4*slog,xbarlog+4*slog, length.out = 50)
yfitlog<-dnorm(xfitlog,mean=xbarlog,sd=slog)
hist(log(cbdata$tripduration), breaks=30, probability = TRUE)
lines(xfitlog,yfitlog,lwd=3,col='green')
```

Looks promising! Now, we will run an experiment by creating 10,000 random samples of the lognormal distribution and recording their medians:

```{r}
med<-array()
for(i in 1:10000){
  m<-rlnorm(n=500)
  med[i]<-median(m)
}

# Plotting the histogram and overlaying a normal fit:
hist(med, breaks=30, probability = TRUE)
medmean<-mean(med)
meds<-sd(med)
xfitmed<-seq(medmean-4*meds,medmean+4*meds, length.out = 50)
yfitmed<-dnorm(xfitmed, mean=medmean, sd=meds)
lines(xfitmed, yfitmed, lwd=3, col='blue')
```

Based on this analysis, the sample median distribution is normal, so we can be confident using a t-test for the median. Since the population variance is not known we will use the t-statistic instead of the z-statistic. We calualate the t-statistic and we will reject the null hypothesis if it is greater than the critical value. Instead of testing the location of the mean, we test on the location of the median:

\[t=\frac{\tilde{x} - \tilde{\mu}}{s/\sqrt{n}}\]

```{r}
med_h0 <- 900   #the hypothesized median
alpha <- 0.05 #significance level

t1<-(xmed-med_h0)/(s/sqrt(n))
t1
qt(1-alpha,df=n-1)
#Or, the p-value:
pt(t1,df = n-1)
```

Because the t-statistic is not greater than the critical value, we do not reject the null hypothesis.

#Problem 2

#Subset the overtime data:
cbd_ot<-subset(cbdata, tripduration>2700)
med_ot<-median(cbd_ot$tripduration)

# H_0: median(OT tripduration) = 7200
# H_a: median(OT tripduration) != 7200
# Since the null hypothesis tests for an equality, we need a two-tailed t-test.

med_ot_h0 <- 7200
alpha <- 0.05
s_ot <- sd(cbd_ot$tripduration)
n_ot <- dim(cbd_ot)[1]
  
t2 <- (med_ot-med_ot_h0)/(s_ot/sqrt(n_ot))
qt(1-alpha/2,df=n_ot-1)

# Since the t-statistic is less than the negative critical value, we reject the null hypothesis that the median 
# for overtime is 2 hours.

#Problem 3

#H_0: no difference in OT mean by gender
#H_a: difference in OT mean by gender

anova <- aov(tripduration ~ gender, data = cbd_ot)
summary(anova)
#The critical value for the test is:
qf(1-alpha,1,214)
#Since the F statistic is not greater than the critical value, we do not reject the null hypothesis